---
title: "Introduction to R"
author: "José Bayoán Santiago Calderón"
date: "2019-06-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to R markdown

What is [Literate
programming](https://en.wikipedia.org/wiki/Literate_programming)?

> "Programs are meant to be read by humans and only incidentally for
> computers to execute." ― Donald Knuth

R markdowns are programs that mix a markup language (i.e., markdown)
with programming languages (e.g., R) to best communicate compared to
having either one in isolation. Many markdown features are implemented
such as **bold**, *italics*, ***bold and italics*** or hyperlinks such
as the previous example. It is easily integrated withing
[RStudio](https://www.rstudio.com/) the flagship integrated development
environment
([IDE](https://en.wikipedia.org/wiki/Integrated_development_environment))
for R.

## Setting up RStudio

-   Global options for RStudio

Many newcomers and advanced users alike can benefit from customizing
their RStudio global options. For example, I personally like a dark dark
mode theme such as Tomorrow Night 80. Many options under the Code menu
can help you adhere to best practices, catch issues, or follow a
standard coding style.

-   The `.Rprofile` in the HOME directory can be used to modify startup
    behavior

For example, my `.Rprofile` file has the following code

```{r cars}
Sys.setenv(TZ = 'EST')
if (interactive()) {
  try(expr = fortunes::fortune(), silent = TRUE)
  }
cat("Welcome, Nosferican.\nLogged at: ", as.character(Sys.time()))
```

which displays an R fortune cookie, my username and current time when I
start a session. The fortune cookie is a random piece of R
history/culture very useful to understand the philosophy behind the
project.

-   The `.Renviron` file can be used to define variables accessible
    during your session

For example, rather than writing my database credentials and making
those avaiable in the code I can specify these as environmental
variables I can access without compromising these

```{r}
db_userid = "[your computing ID]"
db_pwd = "[your database pwd]"
```

## Setting up your environment (Housekeeping)

For the following lesson, we will be using various functionality
available in R packages. The following code will make the namespace
(e.g., functions) available for us to use.

The next block of code will provide all the functionality we will be
using during this lesson

```{r}
library(magrittr)
library(stringr)
library(forcats)
library(lubridate)
library(ggplot2)
```

## Atomic data types

R has six atomic data types:

-   `logical`
-   `integer`
-   `numeric` (alias for double precision)
-   `complex`
-   `character`
-   `raw`

### Logical

```{r}
x <- c(NA, FALSE, TRUE)
names(x = x) <- as.character(x = x)
x
```

`&` Table

```{r &}
outer(X = x, Y = x, FUN = "&")
```

`|` Table

```{r |}
outer(X = x, Y = x, FUN = "|")
```

`xor` Table

```{r xor}
outer(X = x, Y = x, FUN = "xor")
```

Negation

```{r !}
!TRUE
!FALSE
```

Logical interpretation of integers/numeric

```{r}
TRUE & 0 # (TRUE & (0 == 0))
FALSE | -2 # (FALSE | (-2 == 0))
```

### Integer

```{r}
x = 2L
y = 3L
+x # Positive sign
-x # Negative sign
abs(x = x)
x + y
x - y
x * y
x / y
x^y
sqrt(x = x)
x %% y # Modulus
x %/% y # Integer division
log(x = x)
exp(x = x)
factorial(x = x)
```

### Numeric

```{r}
x = 2.5
y = 3.2
```

```{r}
+x # Positive sign
-x # Negative sign
abs(x = x)
x + y
x - y
x * y
x / y
x^y
sqrt(x = x) 
x %% y # Modulus
x %/% y # Integer division
log(x = x) # Log (defaults to natual log)
exp(x = x)
factorial(x = x)
round(x = x, digits = 2L)
floor(x = x)
ceiling(x = x)
```

### Complex

```{r}
x = 2 + 1i
y = 3 + 1i
```

```{r}
+x # Positive sign
-x # Negative sign
abs(x = x)
x + y
x - y
x * y
x / y
x^y
sqrt(x = x)
log(x = x) # Log (defaults to natual log)
exp(x = x)
x == y
is.complex(x = x)
Re(z = x)
Im(z = x)
Mod(z = x)
Arg(z = x)
Conj(z = x)
```

### Character

```{r}
x <- "\"R doesn't like emojis :(, Julia does.\" - Bayoán (史志鼎)"
str_detect(string = x, pattern = "Bayoán")
str_detect(string = x, pattern = "Nosferican")
str_extract(string = x, pattern = "(?<=\" - ).*")
str_sub(string = x, start = 28L, end = 32L)
str_replace(string = x, pattern = "(?<=\" - ).*", "Me")
str_remove(string = x, pattern = "e")
str_replace_all(string = x, pattern = "e", replace = "E")
str_c("This a fine quote.", x, sep = "  ")
```

## Collections

### Arrays

Arrays are collections of homogeneous data types. The most common types
are the one dimensional array (`vector`) and the two dimensional array
(`matrix`).

```{r vector_init}
vector(mode = "logical", length = 2L)
vector(mode = "character", length = 5L)
vector(mode = "numeric", length = 3L)
vector(mode = "integer", length = 4L)
```

```{r vector}
x <- 1:10
str(object = x)
summary(object = x)
table(origininal_object = x)
x[2:5] <- 1L
table(modified_object = x)
x[11] <- NA
typeof(x)
sort(x = x, decreasing = TRUE, na.last = TRUE) # default values TRUE and NA
x[13] <- "A"
x
is.vector(x = x)
```

## Matrices

```{r matrix}
x <- matrix(data = 1:16,
            nrow = 4L,
            ncol = 4L,
            dimnames = list(LETTERS[1:4], letters[1:4]))
x + 2 * x
x %*% x
str(object = x)
summary(object = x)
table(x)
x[3, 2] <- NA
x
```

## Arrays

```{r array}
x <- array(data = 1:3, dim = c(2L, 4L, 3L))
x
str(object = x)
dim(x = x)
```

### Lists

```{r}
x = list(x = 1:5, y = TRUE, z = c("A", "Z"))
str(object = x)
x$z
x$A <- 23L
x
x[1L]
x[[1L]]
x[[4]] <- list(a = 3, b = 2)
x
x$A$b
```

## Structs

Date time objects are structs can wrap an atomic data element for
special functionality

```{r}
x <- now()
year(x = x)
month(x = x)
day(x = x)
hour(x = x)
minute(x = x)
seconds(x = 1)
x %m+% months(x = 1L, abbreviate = TRUE)
```

Named vectors are an example of a struct with more than one atomic data
type element

```{r}
x <- c(a = 1, b = 2)
str(object = x)
names(x = x)
```

One last example includes categorical or factor variables

```{r}
set.seed(seed = 0L)
x <- as_factor(x = sample(x = letters[1:5], size = 100L, replace = TRUE))
str(x)
x[10L]
levels(x = x)
ggplot(mapping = aes(x = x)) +
  geom_bar() +
  labs(title = "Histogram",
       x = "Category",
       y = "Count") +
  theme(plot.title = element_text(hjust = 0.5))
```

# "The 3 Ways of R" or Becoming a Better Code Googler (GoogleR?)

author: Aaron D. Schroeder date: 6/5/2020 autosize: true

# The R Language - Multiple Approaches

# Used many different ways, including

-   **Procedural programming (PP)**
    -   HOW MOST EVERYBODY STARTS, also known as inline programming
        takes a top-down approach. Writing a list of instructions step
        by step.
-   **Object-oriented programming (OOP)**
    -   encapsulating data and behavior into objects. An OOP application
        will use a collection of objects which knows how to perform
        certain actions and how to interact with other elements of the
        application.
-   **Functional programming (FP)**
    -   WHAT MOST PEOPLE GET TO AFTER PROCEDURAL, passing data from
        function to function to function to get a result. Functions are
        treated as data, meaning you can use them as parameters, return
        them, build functions from other functions, etc.

# Like Javascript

### Can be also be used as PP, OOP, or FP

### 'Frameworks' are created to make it easier

-   **React**
-   **Vue.js**
-   **Ember.js**
-   **Meteor.js**

### And many more! BUT, A Single Framework can Lock You In!

### PROBLEM! People are only learning how to do Javascript the "React" or "Vue.js" or ... way - VERY LIMITING - ESPECIALLY FOR USING OTHER PEOPLE'S CODE!

# R 'Frameworks'

### While not exactly "Frameworks", R has the same issue in that different coding "approaches" have emerged

-   **base**
    -   championed purists who like to do things the old-fashioned way
        (reminds them of the good-ol-days before S was renamed to R)
    -   if you know what is going on "under the hood" you can write
        susinct and fast code, but if you don't...
-   **tidy** (tibl, dplyr, stringr, many many more)
    -   championed by the 'followers of the Wickham Way', I'm no going
        to say it's a cult, but...
    -   an "opionated" way to do functional programming that is easier
        to pick up than base
    -   get all data to tidy format, then process in an easy to follow
        chain
    -   way different than base and rather verbose
-   **datatable**
    -   championed by grumpy programmers who value speed and
        sussinctness most
    -   actually just a very fast data structure that can be used with
        tidy or anything else
    -   however, has it's own syntax that can be significantly more
        susinct
-   **hybrid**
    -   combine the speed of data.table with the programming style of
        tidy
    -   more and more packages are appearing to make this possible

# Need to be Bi/Tri Lingual!

# If you are going to be a great R Code Googler (which you want to be!) you need to be able to at least read and understand what is going on in the code samples you find, regardless of the approach used!

# Today's Objectives

# So, for today our objectives are to

-   Get an introduction to very simple versions of the different
    approaches to
    -   reading data into R
    -   susbsetting data (selecting & filtering)
    -   renaming columns
    -   piping
-   Send you off to the interactive DataCamp tutorials to learn more and
    practice
    -   Data Manipulation with dplyr
    -   Data Manipulation with data.table

```{r sample-data}
df1 <- data.frame(a = seq(1,16,by=2), b = LETTERS[1:8], x=month.abb[1:8], y = sample(10:20,10000, replace = TRUE), z=letters[1:8])
data.table::fwrite(df1, "sample_small.csv")
df2 <- data.frame(a = seq(1,16,by=2), b = LETTERS[1:8], x=month.abb[1:8], y = sample(10:20,1000000, replace = TRUE), z=letters[1:8])
data.table::fwrite(df2, "sample_medium.csv")
df3 <- data.frame(a = seq(1,16,by=2), b = LETTERS[1:8], x=month.abb[1:8], y = sample(10:20,10000000, replace = TRUE), z=letters[1:8])
data.table::fwrite(df3, "sample_large.csv")

```

# base

```{r base}
system.time(read.csv <- read.csv("sample_large.csv", stringsAsFactors =
FALSE))

```

# readr (need to assign to new data.frame)

```{r readr}
library(readr)
system.time(read_csv <- read_csv("sample_large.csv", progress = FALSE))

```

# data.table

```{r data.table}
library(data.table)
system.time(fread \<- fread("sample_large.csv"))

```

# more advanced data.table

```{r}
fread("pigeon-racing.csv", drop = 4:11)
fread("pigeon-racing.csv", select = c(1, 2, 3))
fread("pigeon-racing.csv", drop = c("Name", "Arrival"))
fread("pigeon-racing.csv", select = c("Breeder", "Speed"))

```

# A 4th Way!

```{r vroom}
library(vroom)
system.time(vroom <- vroom("sample_large.csv"))
```

```{r}
results_vroom <- microbenchmark(
read.csv = read.csv(file = here("static", "data", "sim-data-large.csv")),
read_csv = read_csv(file = here("static", "data", "sim-data-large.csv")),
vroom = vroom::vroom(file = here("static", "data", "sim-data-large.csv"))
)

```

```{r}
library(microbenchmark)
results <- microbenchmark(
read.csv = read.csv(file = "sample_medium.csv", stringsAsFactors =
FALSE),
read_csv = read_csv(file = "sample_medium.csv", progress = FALSE),
fread = fread(file = "sample_medium.csv"),
vroom = vroom::vroom(file = "sample_medium.csv"),
times = 10
)

library(ggplot2)

autoplot(object = results) +
scale_y\_log10() +
labs(y = "Time [milliseconds], logged")
```

## A data frame, the basic tabular representation of data in R

```{r}
n = c(2, 3, 5) 
s = c("aa", "bb", "cc") 
b = c(TRUE, FALSE, TRUE) 
df = data.frame(n, s, b)       # df is a data frame
df
```

## Create Sample Data Frame

```{r dataset}
df <- data.frame(column_1 = c(1, 2, 3),
                 column_2 = c("A", "B", "C"),
                 column_3 = c("D", "E", "F"),
                 stringsAsFactors = F)
df

#pg <- fread("pigeon-racing.csv")
```

# Subset Rows

## base R

```{r base.rows}
df[df$column_1 == 2,] # have to include df$ before column name and don't forget the comma!
df[df$column_1 == 2,]
```

## dplyr

```{r dplyr.rows}
library(dplyr)
filter(df, column_1 == 2)
filter(df, column_1 == 2)
```

## data.table

```{r data.table.rows}
library(data.table)
setDT(df) # needs to be a data.table first
df[column_1 == 2]
df[column_1 == 2]
```

# Subset Columns

## base R

```{r base.columns}
df[, c("column_2", "column_3")]
```

## dplyr

```{r dplyr.columns}
select(df, c("column_2", "column_3"))
```

## data.table

```{r data.table.columns}
setDT(df) # needs to be a data.table first
df[, .(column_2, column_3)] # the dot . is a shortcut for list()
df[, .(column_2, column_3)]
# or just like a data.frame
df[, c("column_2", "column_3")]
```

# Subset Rows & Columns

## base R

```{r base.rows.and.columns}
df[df$column_1 == 2, c("column_2", "column_3")]
```

## dplyr

```{r dplyr.rows.and.columns}
library(dplyr)

# two steps using the magrittr "pipe"
df %>%
  filter(column_1 == 2) %>%
  select(c("column_2", "column_3")) # don't need to specifcy data.frame in second step
```

## data.table

```{r data.table.rows.and.columns}
setDT(df) # needs to be a data.table first

df[column_1 == 2, .(column_2, column_3)] # the dot . is a shortcut for list()

# or just like a data.frame
df[df$column_1 == 2, c("column_2", "column_3")]
```

# renaming column base

```{r base}

df <- data.frame(column_1 = c(1, 2, 3), column_b = c("A", "B", "C"))
print(df)
names(df)[names(df) == 'column_b'] <- 'column_2'
print(df)
```

# renaming column dplyr

```{r dplyr}
library(dplyr)
df <- data.frame(column_1 = c(1, 2, 3), column_b = c("A", "B", "C"))
print(df)
df <- dplyr::rename(df, "column_2" = column_b)
print(df)
```

# renaming column data.table

```{r data.table}
library(data.table)
df <- data.frame(column_1 = c(1, 2, 3), column_b = c("A", "B", "C"))
print(df)
setnames(df, "column_b", "column_2")
print(df)
```

```{r load-data}
library(data.table)
library(dplyr)
pg <- fread("pigeon-racing.csv")
```

```{r base}
aggregate(Ent ~ Breeder, pg, length)

table(pg$Breeder)

data.frame(table(pg$Breeder))
```

```{r dplyr}
count(pg, Breeder)

#or with magrittr piping
pg %>% count(Breeder)
```

```{r data.table}
#dt[filter, select, group]
pg[, .N, Breeder]
```

```{r data.table-complete-with-piping}
pg[Color == "BBWF", .(Sex, Speed), .(Breeder)][, .N, .(Breeder)][order(-N)]
```

## Nested Option:

```{r}
library(dplyr)
library(data.table)

arrange(
  summarize(
    group_by(
      filter(mtcars, carb > 1),cyl), Avg_mpg = mean(mpg)), desc(Avg_mpg)
  )
```

## Multiple Object Option

```{r}
a <- filter(mtcars, carb > 1) 
b <- group_by(a, cyl) 
c <- summarise(b, Avg_mpg = mean(mpg)) 
d <- arrange(c, desc(Avg_mpg)) 
print(d)
```

## %\>% Option:

```{r}
library(magrittr) 
library(dplyr) 
mtcars %>% 
  filter(carb > 1) %>% 
  group_by(cyl) %>% 
  summarise(Avg_mpg = mean(mpg)) %>% 
  arrange(desc(Avg_mpg))
```

## . is assumed first

```{r}
mtcars %>% 
  filter(carb > 1) %>% 
  lm(mpg ~ cyl + hp, data = .) %>% 
  summary()
```

## data.table piping

```{r make-data}
set.seed(123)
dt <- data.table(a = sample(letters, 1e5, replace = TRUE),
                 b = abs(rnorm(1e5)))

dt
```

## one line

```{r dt-piping}
# adding columns in dt uses :=, $ will still work as well but not for piping
dt[, x := sqrt(b)][, y := b^2][, z := paste0(a , b)]

dt
```

# multiple lines - ewwwww

```{r dt-piping-multiple-lines}
# multiline dt piping is hard to read
dt[, x := sqrt(b)][
  , y := b^2][
    , z := paste0(a , b)]

dt
```

# using magrittr pipes with data.table :-)

```{r dt-with-magrittr-piping}
# Best of both worlds!
dt[, x := sqrt(b)] %>%
  .[, y := b^2] %>%
  .[, z := paste0(a , b)]

dt
```
