---
title: 'Applied R: Intro to plotting w/ ggplot2'
author: "Joanna Schroder"
date: '2023-06-01'
output: html_document
---

## Setup

```{r setup, message = FALSE, warning = FALSE}
# Load Libraries
install.packages("tidyverse")
library(tidyverse)

# Download Datasets
metadata <- read_csv("metadata.csv")
# airquality dataset
```

## Heads up!

-   We will have a coding best-practices lecture next week
-   We will have a mapping visualization session next week
-   We will have a data visualization best-practices lecture later in the summer

## Things to know

-   Three ways of R
    -   Today we will be using tidy syntax
-   Plotting in Rivanna
    -   Rivanna's RStudio version is out of date, so plots to not show up in the plot pane
    -   I'd recommend plotting in an RMarkdown on Rivanna
-   `ggplot2` lingo
    -   "**G**rammar of **G**raphics"
    -   ggplots are built of layers
    -   In tidy syntax, we use `%>%`; in ggplot we use `+` to add layers
    -   `aes()` = "aesthetics"; this is how we map variables/other aesthetics
    -   `geom_` = "geometry"; these are built-in types of mapping for plots (e.g. `geom_bar()` = bar chart, `geom_hist` = histogram)
-   **Generally, it is much easier to work with long data in `ggplot2` than wide**
    -   `dplyr::pivot_longer()` will become your best friend

## There are a lot of `ggplot2` resources online

-   [General R data viz resources](https://www.cs.drexel.edu/~js5365/r-data-viz/other-resources.html)
-   [`ggplot2` function reference](https://ggplot2.tidyverse.org/reference/index.html)
-   [`ggplot2` themes](https://r-charts.com/ggplot2/themes/)
-   [How to create custom `ggplot2` themes (#1 Google Search result written by DSPG Alum Maddie Pickens!)](https://rpubs.com/mclaire19/ggplot2-custom-themes)
-   Google, stackoverflow, GitHub Issues

## Let's start plotting!

### Make a scatter plot

```{r scatter}
# View the dataset to understand its structure (wide or long?), variables, data types 
view(airquality)

# Pipe the dataset into a ggplot() call
airquality %>% 
# Set aesthetic mappings to the x and y axes for all geoms
  ggplot(aes(x = Wind, y = Temp)) +
# Create a point layer
#    geom_point() #+
# Change the aesthetic mappings of just the point layer
   geom_point(aes(color = as.factor(Month))) +
# Add a trendline, alter the method used to create it
  # geom_smooth() #+
   geom_smooth(method = lm) +
# Add statistical information about the trendline, position it in the plot
 #library(ggpmisc)
  stat_poly_eq(use_label(c("R2", "p", "n")),
               label.x = 20,
               label.y = 90)
```

## Make a line graph

```{r line}
# Manipulate the data before piping it into the ggplot() call
airquality %>% filter(Month == 5) %>%
# Set aesthetic mappings to the x and y axes for all geoms
  ggplot(aes(x = Day, y = Temp)) +
# Create a line layer
  geom_line()

#library(lubridate)
# Manipulate the data before piping it into the ggplot() call to create a new date column
#airquality %>% mutate(date = make_date(2000, Month, Day)) %>% 
#  ggplot(aes(x = date, y = Temp)) +
#  geom_line()

#library(lubridate)
# Manipulate the data before piping it into the ggplot() call to create a new date column
airquality %>% mutate(date = make_date(2000, Month, Day)) %>% select(-Month, -Day) %>%
# Pivot the data longer so we can plot all variables at once
  pivot_longer(cols = c("Ozone", "Solar.R", "Wind", "Temp")) %>%
  ggplot(aes(x = date, y = value)) +
  geom_line() +
# Plot all the variables on the same plot as rows
  facet_grid(name ~ .)
```

## Make a bar graph

```{r bar}
# View the dataset to understand its structure (wide or long?), variables, data types 
#view(metadata)
# Pipe the dataset into a ggplot() call
metadata %>% 
# Set aesthetic mapping for just one variable
  ggplot(aes(x = element)) +
# Create a histogram layer and set a statistical transformation for it
  geom_histogram(stat = "count") +
# Change the angle of axis labels
   theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
# Change the color of geoms
   geom_histogram(aes(fill = collection), stat = "count")
# Flip x and y axis
 coord_flip()
```

## Bare-minimum visualization

The plots we've just created are what I would consider "bare-minimum" visualizations. Here's how we take them to the next level:

1.  **Think about our data story.** What do we want to say about our data? Which plot type, color scheme, annotations, etc. are best suited to tell this story? Who is our audience?
2.  **Choose an accessible color palette.** [Read more on accessible colors here](https://www.cs.drexel.edu/~js5365/r-data-viz/guides/accessible-colors.html)
3.  **Pay attention to default text sizes.** The default `ggplot2` text sizes are generally **too small**. Increasing the text size is an easy way to make more accessible, professional looking plots. [Read more on accessible text here](https://www.cs.drexel.edu/~js5365/r-data-viz/guides/accessible-text.html)
4.  **Consider your figure as standing alone.** The best figures are not just a plot, but a plot with a descriptive title, caption, and labels. Empathize with your audience to consider what information they would need to interpret your figure without you there. **Always** include information about the data source(s) in the caption. **Do not** leave variable names unaltered.

## Next level visualization

```{r important, fig.height = 4, fig.width = 6}
# Create a custom color palette
my_pal <- RColorBrewer::brewer.pal(11, "PRGn")[c(2, 9)]

library(ggtext)
library(RColorBrewer)
# Manipulate data before piping it into the ggplot() call
metadata %>% group_by(element, collection) %>% mutate(count = n()) %>% distinct(element, count, collection) %>%
# Set aesthetic mapping for all layers
# Reorder a variable by its value
  ggplot(aes(x = reorder(element, count), y = count, fill = collection)) +
# Create a column layer, set the columns to equal width
  geom_col(position = position_dodge(preserve = "single")) +
# Flip the axes
  coord_flip() +
# Add a default theme before theme alterations
  theme_minimal() +
# Alter the legend position
  theme(legend.position = "bottom",
# Alter the text size
        text = element_text(size = 20)) +
# Add annotations, specify their position and color
  annotate("text", y = 440, x = 16.25, label = "In the Elections Web Archive, subject repeats a mean \n average of 12.8 times per metadata record", color = my_pal[2]) + 
  annotate("text", y = 120, x = 1.8, label = "The Digital Library of the Caribbean metadata contains three \n unique top-level elements: note, abstract, and classification" , color = my_pal[1]) + 
# Modify the titles, axes labels, and caption
  labs(title = "Frequency of Top-Level Elements",
      # subtitle = "Frequency of Top-Level Elements",
       y = "Number of Observations",
       x = "Top-Level Element",
       caption = "Based on 40 random records collected from each repository (80 records total). \n
       Data Sources: https://www.dloc.com/ and \n
       https://www.loc.gov/collections/united-states-elections-web-archive/") +
# Add a custom color palette, alter the names for the legend and variables
  scale_fill_manual(values = my_pal, name = "Collection",
                      labels=c("Digital Library of the Caribbean", "Elections Web Archive"))

# Save our plot
getwd()
ggsave("metadata_exploration-element_comparison_bar.png", plot = last_plot(),
       height = 10, width = 14, units = "in", bg = "white")
```

# Your turn!

```{r your-turn}
# Choose one of the example datasets (or another base R dataset if you know of one)
# Use `ggplot2` to explore and visualize the data

# Create a bare-miniumum visualization or two
# If we have time, take your bare-minimum visualization to the next level

# Report out your data story
```
